---
setwd("C:/Users/jaspe/OneDrive/Bureaublad/Bioinformatics/SFM/TCGA_group2/TCGA_dataset(1)/TCGA_dataset/")---
title: "MSI vs instability tests"
format: html
editor: visual
---

Setup

```{r}
#| message: false
#| warning: false

# Install only if missing
req <- c("readr", "dplyr", "tidyr", "stringr", "ggplot2", "rstatix", "e1071", "purrr")
new <- req[!req %in% installed.packages()[, "Package"]]
if (length(new) > 0) install.packages(new)

library(readr)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(rstatix)
library(e1071)
library(purrr)
```

File paths and load data

```{r}
pred_path <- "prediction_file_crc.csv"
rna_path  <- "tcga_rna_count_data_crc.csv"

pred_raw <- read_csv(pred_path, show_col_types = FALSE)
rna_raw  <- read_csv(rna_path,  show_col_types = FALSE)

cat("Prediction table dimensions:", nrow(pred_raw), "rows x", ncol(pred_raw), "cols\n")
cat("RNA matrix dimensions      :", nrow(rna_raw),  "rows x", ncol(rna_raw),  "cols\n")
```

Quick column preview

```{r}
cat("Prediction columns:\n")
print(colnames(pred_raw))

cat("\nFirst 10 RNA column names:\n")
print(colnames(rna_raw)[1:min(10, ncol(rna_raw))])
```

ID checks + keep overlapping samples

```{r}
# Rename ID columns to avoid confusion
pred <- pred_raw %>% rename(sample_id = `...1`)
rna  <- rna_raw  %>% rename(gene_id   = `...1`)

# Extract IDs
pred_ids <- pred$sample_id
rna_ids  <- colnames(rna)[-1]   # all columns except gene_id

# Duplicate checks
dup_pred <- pred_ids[duplicated(pred_ids)]
dup_rna  <- rna_ids[duplicated(rna_ids)]

cat("Duplicate sample IDs in prediction table:", length(unique(dup_pred)), "\n")
cat("Duplicate sample IDs in RNA matrix     :", length(unique(dup_rna)), "\n")

# Overlap
overlap_ids <- intersect(pred_ids, rna_ids)

cat("Unique prediction IDs:", length(unique(pred_ids)), "\n")
cat("Unique RNA IDs       :", length(unique(rna_ids)), "\n")
cat("Overlapping IDs      :", length(overlap_ids), "\n")

# Keep only overlapping samples
pred_ov <- pred %>% filter(sample_id %in% overlap_ids)

# Subset RNA to overlapping sample columns (keep gene_id + overlap columns)
rna_ov <- rna %>% select(gene_id, all_of(overlap_ids))

pred_ov <- pred_ov %>% slice(match(overlap_ids, sample_id))
stopifnot(identical(pred_ov$sample_id, colnames(rna_ov)[-1]))

#MSI indeterminate
pred_ov <- pred_ov %>%
  mutate(
    msi_status = trimws(as.character(msi_status)),
    msi_status = ifelse(tolower(msi_status) %in% c("indeterminate", "indetermined"), NA, msi_status)
  )

cat("\nAfter filtering:\n")
cat("Prediction rows:", nrow(pred_ov), "\n")
cat("RNA columns    :", ncol(rna_ov), "(includes gene_id)\n")

cat("\nAlignment check (first 10 IDs):\n")
print(head(pred_ov$sample_id, 10))
print(head(colnames(rna_ov)[-1], 10))

cat("\nAligned? ", identical(pred_ov$sample_id, colnames(rna_ov)[-1]), "\n")

```

Missingness checks

```{r}
# Columns we must check (exactly as in your checklist)
cols_check <- c(
  "msi_status",
  "TMB", "TBL", "fraction_genome_altered", "aneuploidy_score",
  "TP53", "KRAS", "BRAF", "APC", "TTN"
)

# Sanity: are all columns present?
missing_cols <- setdiff(cols_check, colnames(pred_ov))
if (length(missing_cols) > 0) {
  stop(paste("These columns are missing in pred_ov:", paste(missing_cols, collapse = ", ")))
}

# % missing per column
missingness_tbl <- pred_ov %>%
  summarise(across(all_of(cols_check), ~ mean(is.na(.) | . == "") * 100)) %>%
  pivot_longer(everything(), names_to = "column", values_to = "pct_missing") %>%
  arrange(desc(pct_missing))

missingness_tbl

n_total <- nrow(pred_ov)
n_msi_missing <- sum(is.na(pred_ov$msi_status) | pred_ov$msi_status == "")
n_msi_present <- n_total - n_msi_missing

cat("Total samples:", n_total, "\n")
cat("MSI missing  :", n_msi_missing, "\n")
cat("MSI present  :", n_msi_present, "\n")
```

Class imbalance checks

```{r}
msi_balance <- pred_ov %>%
  filter(!is.na(msi_status), msi_status != "") %>%
  mutate(msi_status = factor(msi_status, levels = c("MSS", "MSI-L", "MSI-H"))) %>%
  count(msi_status, name = "n") %>%
  mutate(prop = n / sum(n))

msi_balance
```

```{r}
mut_genes <- c("TP53", "KRAS", "BRAF", "APC", "TTN")

mut_balance <- purrr::map_dfr(mut_genes, function(g) {
  pred_ov %>%
    mutate(value = as.character(.data[[g]])) %>%
    mutate(value = trimws(value)) %>%
    filter(!is.na(value), value != "") %>%
    count(value, name = "n") %>%
    mutate(prop = n / sum(n), gene = g) %>%
    select(gene, value, n, prop)
})

mut_balance
```

Distribution checks

```{r}
num_targets <- c("TMB", "TBL", "fraction_genome_altered", "aneuploidy_score")

dist_tbl <- purrr::map_dfr(num_targets, function(v) {

  x_all <- pred_ov[[v]]
  n_total <- length(x_all)
  n_missing <- sum(is.na(x_all) | x_all == "")
  x <- suppressWarnings(as.numeric(x_all))
  x <- x[!is.na(x)]

  if (length(x) == 0) {
    return(tibble(
      variable = v,
      n_total = n_total,
      n_missing = n_missing,
      n_used = 0,
      min = NA_real_, q1 = NA_real_, median = NA_real_, q3 = NA_real_, max = NA_real_,
      range = NA_real_,
      skewness = NA_real_,
      outliers_iqr = NA_integer_
    ))
  }

  q1 <- as.numeric(quantile(x, 0.25, names = FALSE))
  q3 <- as.numeric(quantile(x, 0.75, names = FALSE))
  iqr <- q3 - q1
  low <- q1 - 1.5 * iqr
  high <- q3 + 1.5 * iqr
  out_n <- sum(x < low | x > high)

  tibble(
    variable = v,
    n_total = n_total,
    n_missing = n_missing,
    n_used = length(x),
    min = min(x),
    q1 = q1,
    median = median(x),
    q3 = q3,
    max = max(x),
    range = max(x) - min(x),
    skewness = e1071::skewness(x, type = 2),
    outliers_iqr = out_n
  )
})

dist_tbl
```

```{r}

for (v in c("TMB","TBL","fraction_genome_altered","aneuploidy_score")) {
  print(
    ggplot(pred_ov, aes(x = .data[[v]])) +
      geom_histogram(bins = 30) +
      theme_minimal(base_size = 12) +
      labs(title = paste("Histogram:", v), x = v, y = "Count")
  )
}
```

Dataset for MSI-based tests

```{r}
msi_df <- pred_ov %>%
  filter(!is.na(msi_status), msi_status != "") %>%
  mutate(msi_status = factor(msi_status, levels = c("MSS", "MSI-L", "MSI-H")))

cat("Rows in msi_df:", nrow(msi_df), "\n")
msi_df %>% count(msi_status)
```

**A. MSI vs instability metrics**

MSI vs TMB

```{r}
# 1) Create transformed variable (handles skew; safe if TMB can be 0)
msi_df <- msi_df %>%
  mutate(logTMB = log1p(TMB))

# 2) Quick visual check
ggplot(msi_df, aes(x = msi_status, y = logTMB)) +
  geom_boxplot() +
  theme_minimal(base_size = 12) +
  labs(title = "log1p(TMB) by MSI class", x = "MSI class", y = "log1p(TMB)")

# 3) Fit one-way ANOVA
fit_tmb <- aov(logTMB ~ msi_status, data = msi_df)
summary(fit_tmb)

# 4) Post-hoc pairwise comparisons (Tukey)
TukeyHSD(fit_tmb)

# 5) Diagnostic plots (assumption checks)
par(mfrow = c(1, 2))
plot(fit_tmb, which = 1)  # residuals vs fitted (variance / patterns)
plot(fit_tmb, which = 2)  # Q-Q plot (normality of residuals)
par(mfrow = c(1, 1))
```

MSI vs TBL

```{r}
# 1) Transform (TBL is right-skewed, log helps)
msi_df <- msi_df %>%
  mutate(logTBL = log1p(TBL))

# 2) Boxplot on transformed scale
ggplot(msi_df, aes(x = msi_status, y = logTBL)) +
  geom_boxplot() +
  theme_minimal(base_size = 12) +
  labs(title = "log1p(TBL) by MSI class", x = "MSI class", y = "log1p(TBL)")

# 3) One-way ANOVA
fit_tbl <- aov(logTBL ~ msi_status, data = msi_df)
summary(fit_tbl)

# 4) Tukey post-hoc
TukeyHSD(fit_tbl)

# 5) Diagnostic plots
par(mfrow = c(1, 2))
plot(fit_tbl, which = 1)
plot(fit_tbl, which = 2)
par(mfrow = c(1, 1))
```

```{r}
cor(pred_ov$TBL, pred_ov$fraction_genome_altered, method = "spearman")
cor(pred_ov$TBL, pred_ov$aneuploidy_score, method = "spearman")
cor(pred_ov$TMB, pred_ov$TBL, method = "spearman")
```

MSI vs fraction_genome_altered

```{r}
# 1) Boxplot (raw scale)
ggplot(msi_df, aes(x = msi_status, y = fraction_genome_altered)) +
  geom_boxplot() +
  theme_minimal(base_size = 12) +
  labs(title = "fraction_genome_altered by MSI class",
       x = "MSI class", y = "fraction_genome_altered")

# 2) One-way ANOVA
fit_fga <- aov(fraction_genome_altered ~ msi_status, data = msi_df)
summary(fit_fga)

# 3) Tukey post-hoc
TukeyHSD(fit_fga)

# 4) Diagnostics
par(mfrow = c(1, 2))
plot(fit_fga, which = 1)
plot(fit_fga, which = 2)
par(mfrow = c(1, 1))
```

MSI vs aneuploidy_score

```{r}
# 1) Boxplot (raw scale)
ggplot(msi_df, aes(x = msi_status, y = aneuploidy_score)) +
  geom_boxplot() +
  theme_minimal(base_size = 12) +
  labs(title = "aneuploidy_score by MSI class",
       x = "MSI class", y = "aneuploidy_score")

# 2) One-way ANOVA
fit_aneu <- aov(aneuploidy_score ~ msi_status, data = msi_df)
summary(fit_aneu)

# 3) Tukey post-hoc
TukeyHSD(fit_aneu)

# 4) Diagnostics
par(mfrow = c(1, 2))
plot(fit_aneu, which = 1)
plot(fit_aneu, which = 2)
par(mfrow = c(1, 1))

```

**E. Optional: MSI vs each mutation gene directly – (Jasper)**

Compare:

(Technically overlaps with enrichment section, but you can present gene-by-gene.)

-   MSI class vs TP53

```{r}
# 1) Contingency Table
tp53_table <- table(msi_df$msi_status, msi_df$TP53)
print("TP53 vs MSI Table:")
print(tp53_table)

# 2) Fisher's Exact Test
tp53_test <- fisher.test(tp53_table)
print(tp53_test)

# 3) Visualization
ggplot(msi_df, aes(x = msi_status, fill = TP53)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal(base_size = 12) +
  labs(title = "TP53 Mutation Distribution by MSI Status", 
       x = "MSI Group", y = "Percentage", fill = "TP53 Status") +
  scale_fill_manual(values = c("WT" = "gray70", "SNV" = "firebrick"))
```

-   MSI class vs KRAS

```{r}
# 1) Contingency Table
kras_table <- table(msi_df$msi_status, msi_df$KRAS)
print("KRAS vs MSI Table:")
print(kras_table)

# 2) Fisher's Exact Test
kras_test <- fisher.test(kras_table)
print(kras_test)

# 3) Visualization
ggplot(msi_df, aes(x = msi_status, fill = KRAS)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal(base_size = 12) +
  labs(title = "KRAS Mutation Distribution by MSI Status", 
       x = "MSI Group", y = "Percentage", fill = "KRAS Status") +
  scale_fill_manual(values = c("WT" = "gray70", "SNV" = "darkorange"))
```

-   MSI class vs BRAF

```{r}
# 1) Contingency Table
braf_table <- table(msi_df$msi_status, msi_df$BRAF)
print("BRAF vs MSI Table:")
print(braf_table)

# 2) Fisher's Exact Test
# Using the exact test (no simulation) to get the precise scientific p-value
braf_test <- fisher.test(braf_table)
print(braf_test)

# 3) Visualization
ggplot(msi_df, aes(x = msi_status, fill = BRAF)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal(base_size = 12) +
  labs(
    title = "BRAF Mutation Distribution by MSI Status", 
    x = "MSI Group", 
    y = "Percentage", 
    fill = "BRAF Status"
  ) +
  # WT is Gray, SNV is Purple
  scale_fill_manual(values = c("WT" = "gray70", "SNV" = "purple"))

```

-   MSI class vs APC

    ```{r}
    # 1) Contingency Table
    apc_table <- table(msi_df$msi_status, msi_df$APC)
    print("APC vs MSI Table:")
    print(apc_table)

    # 2) Fisher's Exact Test
    apc_test <- fisher.test(apc_table)
    print(apc_test)

    # 3) Visualization
    ggplot(msi_df, aes(x = msi_status, fill = APC)) +
      geom_bar(position = "fill") +
      scale_y_continuous(labels = scales::percent) +
      theme_minimal(base_size = 12) +
      labs(title = "APC Mutation Distribution by MSI Status", 
           x = "MSI Group", y = "Percentage", fill = "APC Status") +
      scale_fill_manual(values = c("WT" = "gray70", "SNV" = "steelblue"))
    ```

<!-- -->

-   MSI class vs TTN

```{r}
# 1) Contingency Table
ttn_table <- table(msi_df$msi_status, msi_df$TTN)
print("TTN vs MSI Table:")
print(ttn_table)

# 2) Fisher's Exact Test
ttn_test <- fisher.test(ttn_table)
print(ttn_test)

# 3) Visualization
ggplot(msi_df, aes(x = msi_status, fill = TTN)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal(base_size = 12) +
  labs(title = "TTN Mutation Distribution by MSI Status", 
       x = "MSI Group", y = "Percentage", fill = "TTN Status") +
  scale_fill_manual(values = c("WT" = "gray70", "SNV" = "forestgreen"))
```
